Estimating Daily Bicycle Counts in Seattle from Seasonal and Weather Factors

### 2014-12-09 article/article.tex
lit summary table and edits

Bicycling volumn in cities iuseful for practitioners and researchers
to understand safety, travel behavior, and development impacts.
Therefore it is of great interest to investigate the relationship
between bicycle volume and various factors. The goal is to build a
predictive model based on this relationship, so that one can better
estimate and forecast bicycle travel, help bicycle facility planning,
etc. There have been a lot of papers in the last decade related to
this topic (see, for example,
\cite{Griswold:2011aa,Fields:2012aa,Niemeier:1996aa,Nosal:2014aa}, and
reference therein). Regarding data source, survey/census data are more
used to explain influencing factors such as demographic and
socioeconomic factors to mode choice \cite{Parkin:2008aa,Helbich:2014aa} while
self-collected data(automatic or manual) are used in more studies in
recently to track and analyze the count in a long period of
time\cite{Griswold:2011aa,Nosal:2014aa}. Despite a great amount of research,
the methodology is more or less the same: 1) First propose a set of
explanatory variables; 2) Then fit a proper regression model and
justify the result.

According to a recent report by \cite{Bassok:2011aa}, there are eleven
primary indicators, which are time of day \cite{Schwartz:1999aa},
season \cite{Niemeier:1996aa}, population and employment densities
\cite{McCahil:2008aa,Pinjari:2009aa}, mix of use
\cite{Pinjari:2009aa}, bicycle facility type \cite{Hunt:2007aa},
traffic volume \cite{McDonald:2007aa}, rain and temperature
\cite{Niemeier:1996aa,Parkin:2008aa}, income \cite{Turner:1998aa}, and
age \cite{Hunt:2007aa}. This section outlines the key points made in
the literature that are relevant to some of most important variables.
Research has found the variability for counts has a positive
association with high temperature and low precipitation
\cite{Niemeier:1996aa,Parkin:2008aa}. Meanwhile, as suggested by
\cite{Lewin:2011aa} and \cite{Thomas:2009aa}, the effects of
precipitation and temperature on bicycle volumes are nonlinear. For
example, bicycle traffic can decrease in both very cold and very hot
weather as noted in \cite{Richardson:2000aa}. Apart from the usual
temperature and rain variables, \cite{Miranda-Moreno:2011aa} finds
humidity and additional precipitation variables including the presence
of rain in the morning and/or during the previous three hours to be
significant too. Other comparative studies are also available where
bicycle counts are conducted in different cities and different
sensitives to weather are examined \cite{Rose:2011aa}. As for
longitudinal study, \cite{Niemeier:1996aa} finds increased variability
for counts conducted in the later months of the year.
\cite{Jones:2010aa} conclude that morning peak hours from 6 AM to 9 PM
accounts for a consistent 95\% of the total bicycle volumes by hourly
count data. The simple linear regression model has been used in the
many general applications \cite{Jones:2008aa,Jones:2010aa}. Few
exceptions include \cite{Miranda-Moreno:2011aa} which develops a count
model and \cite{Thomas:2009aa} which develops a time-series model.
\cite{Niemeier:1996aa} also uses a Poisson model to statistically
confirm that many of the factors thought to influence cyclists. The
work by \cite{Gallop:2012aa} adopts a similar time-series approach
while incorporating an autoregressive integrated moving average
(ARIMA) analysis.

After reviewing these literature, we found that the bicycle count data
is need to be further analysis with a long-term data and a better
model. How seasonal factors influence bicycle flow needs to be
examined in data that last more than a year. Plus, among the past
researches, almost no literature discuss about the goodness of fit of
their modeling, a model that can better describe and forecast the
bicycle count in longitudinal form is necessary to developed. Models
for count data with better estimation methods are promising.

@@

\begin{centering}
\begin{table}[h]
\begin{scriptsize}
\caption{Variables included in model specification}
\begin{tabularx}{0.5\textwidth}{>{\raggedright\arraybackslash}p{1.6cm}>{\raggedright\arraybackslash}X}
\toprule
Variable & Description \\
\midrule
Count* & Number of bicycles per day \\
Daylight  & Length of daylight in hours \\
UW & Was the University of Washington in session? \\
Holiday & Was the day a holiday as recognized by UW? \\
Max temp & Maximum temperature for the day \\
Max precip & Daily maximum inches of precipitation in any hour \\
Sat--Fri & Day of the week dummy variable (relative to Sunday) \\
Day \# & Sequentially numbered day of study \\
\bottomrule
\multicolumn{2}{r}{* Dependent variable}
\end{tabularx}

>>>>

Bicycling volume in cities is useful for practitioners and researchers
to understand safety, travel behavior, and development impacts.
Therefore the relationship between bicycle volume and various factors,
with the goal to build a predictive model based on this relationship,
has been of great interest to researchers over the last decade (e.g.
\cite{Griswold:2011aa,Fields:2012aa,Niemeier:1996aa,Nosal:2014aa}).

To address these questions, researchers have used a variety of
approaches using different data sources. Studies employing survey or
census data are more often used to explain influences such as
demographic and socioeconomic factors on mode choice
\parencite{Parkin:2008aa,Helbich:2014aa}. In other studies, collected
count data (from automated counters or manual counts) have more been
used recently to track and analyze counts over longer periods of
time \parencite{Griswold:2011aa,Nosal:2014aa}. These counts-based
approaches have largely followed a similar approach of first proposing
a set of explanatory variables, fitting some form of regression model,
and then interpreting / justifying the results vis-\`{a}-vis the
guiding theory.

A literature review accompanying a recent report by
\textcite{Bassok:2011aa} identified eleven primary indicators, which are
time of day \parencite{Schwartz:1999aa}, season \parencite{Niemeier:1996aa},
population and employment densities
\parencite{McCahil:2008aa,Pinjari:2009aa}, mix of use
\parencite{Pinjari:2009aa}, bicycle facility type \parencite{Hunt:2007aa},
traffic volume \parencite{McDonald:2007aa}, rain and temperature
\parencite{Niemeier:1996aa,Parkin:2008aa}, income \parencite{Turner:1998aa}, and
age \parencite{Hunt:2007aa}. This section outlines the key points made in
the literature that are relevant to some of most important variables.

Research has found the variability for counts has a positive
association with high temperature and low precipitation. \parencite{Niemeier:1996aa,Parkin:2008aa}. Meanwhile, as suggested by
\textcite{Lewin:2011aa} and \textcite{Thomas:2009aa}, the effects of
precipitation and temperature on bicycle volumes are nonlinear. For
example, bicycle traffic can decrease in both very cold and very hot
weather as noted by \textcite{Richardson:2000aa}. Apart from the usual
temperature and rain variables, \textcite{Miranda-Moreno:2011aa} finds
humidity and additional precipitation variables including the presence
of rain in the morning and/or during the previous three hours to be
significant too. Other comparative studies are also available where
bicycle counts are conducted in different cities and different
sensitives to weather are examined \parencite{Rose:2011aa}. As for
longitudinal studies, \textcite{Niemeier:1996aa} finds increased variability
for counts conducted in the later months of the year.
\textcite{Jones:2010aa} conclude that morning peak hours from 6 AM to 9 PM
accounts for a consistent 95\% of the total bicycle volumes by hourly
count data. The simple linear regression model has been used in
several applications \cite{Jones:2008aa,Jones:2010aa}. Other modeling approaches include \cite{Miranda-Moreno:2011aa} which develops a count
model and \cite{Thomas:2009aa} which develops a time-series model.
\cite{Niemeier:1996aa} also uses a Poisson model to statistically
confirm  many of the factors thought to influence cyclists. The
work by \cite{Gallop:2012aa} adopts a similar time-series approach
while incorporating an autoregressive integrated moving average
(ARIMA) analysis. A summary of findings from the literature are
presented in table \ref{tb:lit}.

These literature together suggest an opportunity for further model
development base around long-term automated counts utilizing
appropriate statistical methods. How seasonal factors influence
bicycle flow needs to be examined in data that last more than a year.
One limitation present in much of the past literature is that few
discuss goodness of fit of their modeling. Further a model that can
better describe and forecast the bicycle count in longitudinal form is
necessary to developed. Models for count data with better estimation
methods offer some promise.

\newfixedcaption{\freetabcaption}{table}
\begin{centering}
\begin{scriptsize}
\freetabcaption{Summary of findings from literature review}
\begin{ctabular}{p{.6in}p{1.25in}p{1.25in}}
  \toprule
  Source & Variable(s) identified & Methods \\
  \midrule
  \cite{Fields:2012aa} & Average daily temperature (F); Total weekly precipitation (inches) & Identify patterns through scatter plots; No explicit model is established. \\
  \cite{Gallop:2012aa} & Temperature, Relative humidity, wind speed, clearness,fog, precipitation & Use ARIMA to account for serial correlation patterns \\
  \cite{Griswold:2011aa} & Nearby population and employment density, proximity to downtown/freeway, age, education level, income, et. & Log linear ordinary least squares regression is used to estimate a bicycle count model \\
  \cite{Helbich:2014aa} & Daily maiximum air temperature (C), daily average wind speed (m/s), daily precipitation(mm) & Place-specific associations of weather conditions are explored through geographically weighted logit models \\
  \cite{Hunt:2007aa} & Descriptive variables indicating lane use, secured parking, level of experience, etc. & Logistic model of cycling-related choices \\
  \cite{Jones:2010aa} & Length of bicycle network, employment density, population density & Standard ordinary least squares regression \\
  \cite{Lewin:2011aa} & Max temp (F), rain flag, snow flag, weekend flag, over 90 flag & Standard linear regression model \\
  \cite{McCahil:2008aa} & logarithmic choice measure, population density, worker density & A new space syntax theory is used to evaluate and predict the bicycle volume throughout a network \\
  \cite{Miranda-Moreno:2011aa} & Temperature (C), humidity (\%), rain presence (0/1), rain in prev. 3hrs (0/1), warm \& humid (0/1), morning rain (0/1) & Both log-linear model and negative binomial model are tested \\
  \cite{Niemeier:1996aa} & Morning flag, rain flag, high temp flag, location variable, season variable & A Poisson bicycle count model is assumed and fitted \\
  \cite{Nosal:2014aa} & Temperature (F), humidity (\%), rain flag, rain prev. 3hrs, am rain, pm rain & The relationship  is analyzed using a log-linear regression model \\
  \cite{Parkin:2008aa} & Gender, car ownership, hilliness, off-road routes proportion & A logistic regression model of relevant socio-economic and physical variables is estimated. \\
  \cite{Pinjari:2009aa} & Household density, employment density, fraction of commercial land area, fraction of senior, fraction of race population, etc. & The model system takes the form of a joint mixed Multinomial Logit–Multiple Discrete-Continuous Extreme Value (MNL–MDCEV) structure \\
  \cite{Rose:2011aa} & Temerature(C), rainfall(mm), holiday flag, school seanson flag, day of the week & Weather and other effects examined using an aggregate model of daily ridership \\
  \cite{Thomas:2009aa} & Temperature(C), sunshine(h), precipitation(mm), wind force(m/s), cycle path use & A bi-level structure is developed with the upper level being a log-linear model and the lower level being a linear model \\
  \bottomrule
\label{tb:lit}
\end{ctabular}
\end{scriptsize}
\end{centering}

@@

\begin{centering}
\begin{table}[ht!]
\begin{scriptsize}
\caption{Variables included in model specification}
\begin{tabularx}{0.5\textwidth}{>{\raggedright\arraybackslash}p{1.6cm}>{\raggedright\arraybackslash}X}
\toprule
Variable & Description \\
\midrule
Count* & Number of bicycles per day \\
Daylight  & Length of daylight in hours \\
UW & Was the University of Washington in session? \\
Holiday & Was the day a holiday as recognized by UW? \\
Max temp & Maximum temperature for the day \\
Max precip & Daily maximum inches of precipitation in any hour \\
Sat--Fri & Day of the week dummy variable (relative to Sunday) \\
Day \# & Sequentially numbered day of study \\
\bottomrule
\multicolumn{2}{r}{* Dependent variable}
\end{tabularx}
